---
title: "Paul's Notes · Week 3: Epidemiologic Integration and Exploration"
author: "Paul Skeffington"
output:
  html_document:
    toc: true
    toc_depth: 2
    css: ../06_RESOURCES/Presentation/reading.css
---

Combine synthetic demographic, laboratory, and survey tables, then explore an independently simulated biomarker cohort. These invented records demonstrate epidemiologic data handling. Serum albumin and creatinine are measured biomarkers; a coded kidney-health response is self-reported information. **Neither an invented value nor a self-report establishes a clinical diagnosis.**

**Study time:** 120–160 minutes. **Prerequisites:** Weeks 1–2 data frames, keys, factors, missingness, distributions, and units. **Dependencies:** base R, dplyr >= 1.1.0, and ggplot2. Run blocks in order in a clean session.

[Previous: Week 2](Week_2_Group_Work_Narrative_Walkthrough.md) · [Weekly index](README.md) · [Topic companion](../02_Lecture_Notes/Week_3_Data_Visualization_and_Analytics_Lecture_Notes.md) · [Next: Week 4](Week_4_Bash_and_Reproducible_Workflows_Study_Guide.md)

## Learning objectives

Import typed tables; read a data dictionary; validate unique keys and join cardinality; audit matched/unmatched records; distinguish absent measurements from coded responses; define an analytic sample; draw distribution and association views; fit and diagnose an exploratory linear model; create checked plotting functions and multi-panel figures.

## Scientific motivation and dictionary

Health studies often store demographic information, assays, and questionnaire responses separately. Joining them changes which records are visible and which fields are missing. Decide what each row means before combining tables; retaining a row does not guarantee it is eligible for every analysis.

| Field | Definition and type |
| --- | --- |
| Subject_ID | Character participant key, retained exactly across files |
| Age | Numeric years in invented adult participants |
| Sex | Fictional recorded female/male category; not a comprehensive representation of sex or gender |
| Albumin | Synthetic serum albumin concentration, g/dL |
| Creatinine | Synthetic serum creatinine concentration, mg/dL |
| Weak_Kidney | Toy survey code: 1 reported yes, 2 reported no, 7 refused, 9 unsure; NA is an unavailable response |

The survey label abbreviates a fictional self-report item and its codes belong only to this example. It is not a validated disease definition or an institutional/public survey codebook. Laboratory definitions and units can be checked in [albumin background](https://medlineplus.gov/lab-tests/albumin-blood-test/) and [creatinine background](https://medlineplus.gov/lab-tests/creatinine-test/). The values below are synthetic, with no diagnostic classification.

## 1. Dependency check and structured import

```{r dependencies}
for (package in c("dplyr", "ggplot2")) {
  if (!requireNamespace(package, quietly = TRUE)) stop("Install in your own R library: ", package)
}
if (utils::packageVersion("dplyr") < "1.1.0") stop("Need dplyr >= 1.1.0 for relationship checks")
```

**Expected:** no output. Install missing packages separately; the lesson never silently changes your package library.

```{r synthetic-file-import}
demographics_input <- data.frame(
  Subject_ID = sprintf("E%03d", 701:706),
  Age = c(31, 46, 54, 63, 40, 72),
  Sex = c("Female", "Male", "Female", "Male", "Male", "Female")
)
lab_input <- data.frame(
  Subject_ID = sprintf("E%03d", c(701, 702, 704, 706, 799)),
  Albumin = c(4.1, 4.3, 3.9, 4.6, 4.2),
  Creatinine = c(1.1, 0.9, 1.3, 0.8, 1.0)
)
survey_input <- data.frame(Subject_ID = demographics_input$Subject_ID,
                           Weak_Kidney = c(1L, 2L, 7L, 9L, NA_integer_, 2L))
practice_folder <- tempfile("epidemiology_")
dir.create(practice_folder)
inputs <- list(demographics = demographics_input, laboratory = lab_input, survey = survey_input)
for (name in names(inputs)) {
  write.csv(inputs[[name]], file.path(practice_folder, paste0(name, ".csv")), row.names = FALSE, na = "NA")
}
file_paths <- list.files(practice_folder, pattern = "\\.csv$", full.names = TRUE)
stopifnot(length(file_paths) == 3L)
# Explicit types keep identifiers from being coerced or losing leading zeros.
demographics <- read.csv(file.path(practice_folder, "demographics.csv"),
  colClasses = c("character", "numeric", "character"), na.strings = "NA")
laboratory <- read.csv(file.path(practice_folder, "laboratory.csv"),
  colClasses = c("character", "numeric", "numeric"), na.strings = "NA")
survey <- read.csv(file.path(practice_folder, "survey.csv"),
  colClasses = c("character", "integer"), na.strings = "NA")
unlink(practice_folder, recursive = TRUE)
stopifnot(identical(names(demographics), c("Subject_ID", "Age", "Sex")))
stopifnot(identical(names(laboratory), c("Subject_ID", "Albumin", "Creatinine")))
stopifnot(identical(names(survey), c("Subject_ID", "Weak_Kidney")))
stopifnot(is.numeric(laboratory$Albumin), is.numeric(laboratory$Creatinine))
print(lapply(list(demographics, laboratory, survey), dim))
```

**Expected:** demographic 6×3, laboratory 5×3, and survey 6×2 tables. E799 has laboratory data but no demographic row. Temporary files demonstrate import without external data; cleanup removes only the directory created here. In a real study, keep source files unchanged and retain their dictionary and import decisions.

## 2. Unique keys and join choices

Missing or repeated keys violate an intended one-row-per-person contract. Repeated visits require a compound key; they are not necessarily data errors.

```{r join-audits}
for (table in list(demographics, laboratory, survey)) {
  stopifnot(is.character(table$Subject_ID), !anyNA(table$Subject_ID),
            all(nzchar(table$Subject_ID)), !anyDuplicated(table$Subject_ID))
}
matched <- dplyr::inner_join(demographics, laboratory, by = "Subject_ID", relationship = "one-to-one")
retained <- dplyr::left_join(demographics, laboratory, by = "Subject_ID", relationship = "one-to-one")
all_keys <- dplyr::full_join(demographics, laboratory, by = "Subject_ID", relationship = "one-to-one")
unmatched_demographic <- dplyr::anti_join(demographics, laboratory, by = "Subject_ID")
unmatched_lab <- dplyr::anti_join(laboratory, demographics, by = "Subject_ID")
print(retained)
stopifnot(nrow(matched) == 4L, nrow(retained) == 6L, nrow(all_keys) == 7L)
stopifnot(identical(unmatched_demographic$Subject_ID, c("E703", "E705")))
stopifnot(identical(unmatched_lab$Subject_ID, "E799"))
```

**Expected:** inner = four matched participants; left = all six demographic participants; full = seven distinct keys. E703 and E705 have absent laboratory fields in the left result; E799 is unmatched on the laboratory side. A right join retains all right-input rows; reversing the input order and using a left join gives the equivalent retention choice. Join type expresses an analysis population, not just syntax.

```{r duplicate-key-demonstration}
repeated_lab <- rbind(laboratory, laboratory[1, ])
join_error <- tryCatch(dplyr::left_join(demographics, repeated_lab,
  by = "Subject_ID", relationship = "one-to-one"), error = function(e) conditionMessage(e))
stopifnot(is.character(join_error), length(join_error) == 1L)
expanded <- dplyr::left_join(demographics, repeated_lab, by = "Subject_ID", relationship = "one-to-many")
stopifnot(nrow(expanded) == 7L)
```

**Expected:** the one-to-one join rejects the duplicate; one-to-many yields seven rows because E701 matches twice. Resolve whether this represents an accidental duplicate, a second specimen, or a repeat visit before summarizing. Arbitrary `distinct()` can hide conflicting measurements. Many-to-many matches may multiply rows further. `na_matches = "never"` prevents missing-to-missing matches when that is appropriate; this example rejects missing keys earlier.

## 3. Decode self-report without creating a diagnosis

```{r survey-codes}
health_data <- dplyr::left_join(retained, survey, by = "Subject_ID", relationship = "one-to-one")
known_codes <- c(1L, 2L, 7L, 9L)
stopifnot(all(health_data$Weak_Kidney[!is.na(health_data$Weak_Kidney)] %in% known_codes))
health_data$Kidney_Response <- factor(health_data$Weak_Kidney, levels = known_codes,
  labels = c("Reported yes", "Reported no", "Refused", "Unsure"))
health_data$Lab_Available <- !is.na(health_data$Creatinine)
print(table(health_data$Kidney_Response, useNA = "ifany"))
stopifnot(sum(health_data$Weak_Kidney %in% c(1L, 2L)) == 3L)
stopifnot(sum(is.na(health_data$Kidney_Response)) == 1L)
stopifnot(sum(!health_data$Lab_Available) == 2L)
```

**Expected:** yes 1, no 2, refused 1, unsure 1, missing 1 among six participants. There are four observed creatinine values. Refusal and uncertainty are observed responses; NA alone does not reveal a reason. Missing lab fields caused by no matched record differ from an unavailable measurement inside a matched record. Do not average survey codes, or label a yes response as diagnosed kidney disease.

## 4. Construct an exploratory cohort and explicit analytic sample

Use a separate larger cohort to make plotting and model mechanics visible. This generator creates an age–creatinine association by design; it does not estimate an actual epidemiologic relationship.

```{r cohort-simulation}
set.seed(90417)
n <- 60L
cohort <- data.frame(
  Subject_ID = sprintf("C%03d", seq_len(n)),
  Age = rep(seq(25, 70, by = 5), each = 6),
  Sex = factor(rep(c("Female", "Male"), length.out = n), levels = c("Female", "Male"))
)
cohort$Albumin <- 4.3 - 0.007 * (cohort$Age - 50) + rnorm(n, 0, 0.15)
cohort$Creatinine <- 0.85 + 0.008 * (cohort$Age - 50) + rnorm(n, 0, 0.10)
stopifnot(nrow(cohort) == 60L, !anyDuplicated(cohort$Subject_ID))
stopifnot(all(cohort$Albumin > 0), all(cohort$Creatinine > 0))
cohort$Creatinine[c(3, 42)] <- NA_real_
cohort$Age[7] <- NA_real_
# This model requires finite age AND creatinine; unrelated fields do not define exclusion.
included <- is.finite(cohort$Age) & is.finite(cohort$Creatinine)
analytic <- cohort[included, , drop = FALSE]
print(c(total = nrow(cohort), included = sum(included), excluded = sum(!included)))
stopifnot(nrow(analytic) == 57L, sum(is.na(cohort$Creatinine)) == 2L)
print(summary(analytic[c("Age", "Albumin", "Creatinine")]))
```

**Expected:** 60 simulated participants, 57 included, three excluded for the specified model. Age ranges from 25–70 years among observed participants. Albumin is in g/dL, creatinine in mg/dL. The model's generating slope is 0.008 mg/dL per year, with independent normal noise. Sex has no programmed effect. Rounding or differences in printing do not change the input contract.

Complete cases are specific to the analysis fields. A missing unrelated survey response should not automatically exclude a participant from an age–creatinine plot. Available-case analysis can still be biased if missingness relates to outcomes; recording counts is necessary but does not resolve that problem.

## 5. Frequency, density, cumulative, violin, and box views

```{r distribution-views}
creatinine_values <- cohort$Creatinine[is.finite(cohort$Creatinine)]
creatinine_hist <- hist(creatinine_values, breaks = 8, probability = TRUE,
  main = "Synthetic creatinine: 58 measured of 60 participants",
  xlab = "Creatinine (mg/dL)", ylab = "Density (per mg/dL)", col = "grey85")
lines(density(creatinine_values), col = "navy", lwd = 2)
stopifnot(sum(creatinine_hist$counts) == 58L)
creatinine_ecdf <- ecdf(creatinine_values)
plot(creatinine_ecdf, xlab = "Creatinine (mg/dL)", ylab = "Fraction at or below value",
     main = "Empirical cumulative distribution", verticals = TRUE)
print(creatinine_ecdf(1))
stopifnot(creatinine_ecdf(1) == mean(creatinine_values <= 1))
observed_lab <- cohort[is.finite(cohort$Creatinine), ]
p_distribution <- ggplot2::ggplot(observed_lab, ggplot2::aes(Sex, Creatinine)) +
  ggplot2::geom_violin(fill = "lightblue", trim = TRUE) +
  ggplot2::geom_boxplot(width = 0.15, outlier.shape = NA) +
  ggplot2::geom_point(position = ggplot2::position_jitter(width = 0.06, seed = 12), alpha = 0.5) +
  ggplot2::labs(title = "A: synthetic distributions (n = 58)", x = "Recorded sex category",
                y = "Creatinine (mg/dL)") + ggplot2::theme_minimal()
print(p_distribution)
```

**Expected:** histogram/density, ECDF, and violin/box views from 58 measured participants. The ECDF at 1 mg/dL is the measured proportion at or below that illustrative value; it is not a diagnostic proportion. Histogram areas represent proportions when scaled to density. Curve height is not probability at a point. Violin width reflects a smoothed distribution under its scaling rule, not automatically sample count. The box shows median and middle half; whiskers/outliers follow the plotting rule, not disease boundaries. Points make individual observations visible.

## 6. Exploratory regression and units

```{r regression-fit}
creatinine_model <- lm(Creatinine ~ Age, data = analytic, na.action = na.fail)
print(round(coef(creatinine_model), 4))
print(summary(creatinine_model)$r.squared)
stopifnot(nobs(creatinine_model) == 57L)
stopifnot(identical(names(coef(creatinine_model)), c("(Intercept)", "Age")))
stopifnot(abs(coef(creatinine_model)["Age"] - 0.008) < 0.004)
p_relationship <- ggplot2::ggplot(analytic, ggplot2::aes(Age, Creatinine)) +
  ggplot2::geom_point() + ggplot2::geom_smooth(method = "lm", formula = y ~ x, se = TRUE) +
  ggplot2::labs(title = "B: synthetic age association (n = 57)", x = "Age (years)",
                y = "Creatinine (mg/dL)") + ggplot2::theme_minimal()
print(p_relationship)
```

**Expected:** 57 modeled observations and a slope near 0.008, within 0.004 for this seeded example. Interpret the fitted slope as the model's mean creatinine difference in mg/dL per additional year. The intercept predicts age zero, outside this adult sample. R-squared describes explained sample variation, not causal strength or out-of-sample accuracy.

The line estimates a mean trend; its confidence band is not an individual prediction interval. In observational studies, confounding, selection, measurement error, and reverse causation may contribute to associations. These synthetic data encode the association explicitly and provide no evidence about patients.

## 7. Model diagnostics and study assumptions

```{r diagnostic-plots}
plot(fitted(creatinine_model), resid(creatinine_model), pch = 19,
  xlab = "Fitted creatinine (mg/dL)", ylab = "Residual (mg/dL)", main = "Residuals versus fitted")
abline(h = 0, lty = 2)
qqnorm(resid(creatinine_model), main = "Normal Q-Q plot of creatinine residuals",
       xlab = "Theoretical normal quantile", ylab = "Residual (mg/dL)")
qqline(resid(creatinine_model), col = "firebrick")
stopifnot(length(resid(creatinine_model)) == 57L)
```

**Expected:** two diagnostic views. Residuals are observed minus fitted values in mg/dL. Curvature suggests a mean specification issue; widening spread suggests nonconstant variance. The Q-Q horizontal axis gives theoretical normal quantiles and the vertical axis ordered residuals. Normal errors support conventional exact small-sample inference; independence comes from study design and cannot be established by a pleasing residual plot. Repeated visits, influential points, and unmeasured variables need additional checks. Diagnostics flag concerns rather than certifying a model.

## 8. Reusable plotting and a multi-panel figure

```{r checked-plot-function}
plot_biomarker <- function(tbl, variable, unit) {
  if (!is.data.frame(tbl)) stop("tbl must be a data frame")
  if (!is.character(variable) || length(variable) != 1L || is.na(variable) || !nzchar(variable)) {
    stop("variable must be one column name")
  }
  if (!is.character(unit) || length(unit) != 1L || is.na(unit) || !nzchar(unit)) stop("unit must be one label")
  if (!variable %in% names(tbl) || !is.numeric(tbl[[variable]])) stop("Need the requested numeric column")
  valid <- is.finite(tbl[[variable]])
  if (sum(valid) < 3L) stop("Need three finite measurements")
  graph <- ggplot2::ggplot(tbl[valid, , drop = FALSE], ggplot2::aes(x = .data[[variable]])) +
    ggplot2::geom_histogram(bins = 8, fill = "steelblue", colour = "white") +
    ggplot2::labs(x = paste0(variable, " (", unit, ")"), y = "Measured participants") + ggplot2::theme_minimal()
  list(plot = graph, measured = sum(valid), excluded = sum(!valid))
}
units <- c(Albumin = "g/dL", Creatinine = "mg/dL")
plot_records <- lapply(names(units), function(variable) plot_biomarker(cohort, variable, units[[variable]]))
for (record in plot_records) print(record$plot)
stopifnot(identical(vapply(plot_records, function(x) x$measured, integer(1)), c(60L, 58L)))
stopifnot(identical(vapply(plot_records, function(x) x$excluded, integer(1)), c(0L, 2L)))
invalid_plot <- tryCatch(plot_biomarker(cohort, "missing_column", "mg/dL"), error = function(e) conditionMessage(e))
stopifnot(identical(invalid_plot, "Need the requested numeric column"))
# grid ships with R; each ggplot is drawn in its own panel without ggpubr.
grid::grid.newpage()
grid::pushViewport(grid::viewport(layout = grid::grid.layout(1, 2)))
print(p_distribution, vp = grid::viewport(layout.pos.row = 1, layout.pos.col = 1))
print(p_relationship, vp = grid::viewport(layout.pos.row = 1, layout.pos.col = 2))
grid::popViewport()
```

**Expected:** two reusable histogram records with measured/excluded counts 60/0 and 58/2, a caught invalid-column error, and a two-panel figure. Panel A uses 58 measured creatinine values; Panel B additionally requires known age and uses 57. Captions must explain this difference. A function should return its audit counts as well as its plot, and validate schema and labels rather than silently accepting arbitrary columns.

## 9. Record the analysis decisions

```{r reporting-record}
analysis_record <- list(row_unit = "one synthetic participant", total = nrow(cohort),
  modeled = nobs(creatinine_model), excluded = sum(!included),
  slope_mg_dL_per_year = unname(coef(creatinine_model)["Age"]), seed = 90417L,
  r_version = R.version.string, dplyr_version = as.character(utils::packageVersion("dplyr")),
  ggplot2_version = as.character(utils::packageVersion("ggplot2")))
print(analysis_record)
stopifnot(analysis_record$modeled == 57L, analysis_record$excluded == 3L)
```

**Expected:** the computed slope, seed, counts, and installed versions. Report the dictionary, joining population, unmatched IDs, inclusion fields, denominators, formula, units, diagnostics, and limitations. Generate numerical claims from analysis objects rather than copying values from a previous run.

## Common mistakes and debugging

| Symptom | Check and repair |
| --- | --- |
| Join adds unexpected rows | Audit key uniqueness and relationship on both sides |
| Many fields become NA | Check key types, whitespace, and unmatched IDs before calling it nonresponse |
| Survey code treated as a measurement | Decode the toy dictionary; keep refusals/unknowns distinct |
| Model silently loses rows | Construct its analytic sample explicitly and use `na.action = na.fail` |
| Violin looks convincing with tiny n | Inspect points and counts; smoothing is display, not evidence |
| Function drops data invisibly | Return measured/excluded counts and check finite values |
| Panel denominators differ | Explain each panel's inclusion fields in its caption |

## Independent practice

1. Add an unmatched subject and a repeated specimen key. Predict join sizes and demonstrate that an inappropriate one-to-one declaration fails.
2. Insert an undocumented survey code; reject it before factor conversion. Distinguish refusal, uncertainty, and an absent record.
3. Make noise increase with age in a new synthetic cohort. Predict and inspect the residual pattern without making clinical claims.
4. Extend the function to reject a unit mismatch using a supplied data dictionary; test text, all-missing, and nonexistent columns.
5. Create a third panel for albumin and state its separate measured count. Write a caption that explains why panel samples differ.

## Ready to move on

| Evidence | Mastery check |
| --- | --- |
| Input contracts | Typed CSV imports, exact schema, keys, dictionary, and units |
| Joins | Explain inner 4, left 6, full 7; detect duplicate-induced multiplication |
| Survey | Account for yes/no/refused/unsure/missing without diagnosing disease |
| Analytic sample | Explain 60 total, 58 measured creatinine, and 57 model pairs |
| Figures/model | Interpret violin, box, ECDF, scatter, slope units, residuals, and Q-Q axes |
| Reuse/reporting | Return plot/count records and reproduce a panel with a defensible caption |

Rerun from a clean session and keep the dictionary, code, figures, counts, and method record.

## Further learning

[Previous: Week 2](Week_2_Group_Work_Narrative_Walkthrough.md) · [Next: Week 4](Week_4_Bash_and_Reproducible_Workflows_Study_Guide.md) · [R function sheet](../06_RESOURCES/R/EASY_FUNCTION_SHEET.md) · [Paul's Notes](../README.md)

References: [dplyr joins](https://dplyr.tidyverse.org/reference/mutate-joins.html), [ggplot2 violins](https://ggplot2.tidyverse.org/reference/geom_violin.html), [R linear models](https://stat.ethz.ch/R-manual/R-devel/library/stats/html/lm.html), and installed help `?density`, `?ecdf`, `?grid.layout`.
