---
title: "Paul's Notes · Week 1: R for Biomedical Observations"
author: "Paul Skeffington"
output:
  html_document:
    toc: true
    toc_depth: 2
    css: ../06_RESOURCES/Presentation/reading.css
---

Learn R by building a small synthetic expression study. A sample is a collected specimen; a gene is a biological feature; an expression value describes a gene in a sample. Here the gene labels and values are invented. **Expression is in arbitrary simulated units, not raw sequencing counts, normalized RNA-seq output, or patient evidence.**

**Study time:** 90–120 minutes. **Prerequisites:** basic arithmetic and an installed R console or script editor. **Dependencies:** base R only. Run the named blocks in order in a fresh session. The reading edition shows code without executing it.

[Weekly index](README.md) · [Topic companion](../02_Lecture_Notes/Week_1_Introduction_to_R_Lecture_Notes.md) · [Next: Week 2](Week_2_Group_Work_Narrative_Walkthrough.md)

## Learning objectives

Construct and inspect vectors, matrices, lists, factors, and data frames; explain coercion; select observations with logical operators; count missingness; trace a conditional, loop, and function; import/export a table; validate a labeled histogram.

## Scientific motivation and a reproducible session

Biomedical analysis depends on matching values to the correct specimen and measurement definition. A calculation can run successfully on the wrong column. Begin with the row unit, identifier, units, and expected structure rather than a statistical command.

The console evaluates one expression at a time. A saved `.R` script records a repeatable sequence; `.Rmd` combines explanations with executable chunks. Do not type the console's `>` prompt into a script. Use `getwd()` to inspect the working directory: relative paths resolve there. An absolute path identifies a location from the filesystem root. Prefer a project directory and documented relative paths over repeated `setwd()` calls tied to one computer.

Run `R.version.string` to record your version. Save your practice code, restart R, and run it from the first line. Temporary files in this lesson keep existing data unchanged.

## 1. Assignment, vectors, and coercion

Assignment with `<-` names a value. Names are case-sensitive. An atomic vector has one shared storage type; a comment beginning with `#` explains the code without executing it.

```{r expression-values}
expression_values <- c(24, 31, NA_real_, 28, 37, 30)
sample_ids <- sprintf("S%02d", seq_along(expression_values))
print(length(expression_values))
print(typeof(expression_values))
print(class(expression_values))
stopifnot(length(expression_values) == 6L, is.double(expression_values))
```

**Expected:** `6`, `"double"`, and `"numeric"`. `c()` combines values; `NA_real_` marks an unavailable numeric value. Missing expression is not zero expression. Storage type concerns representation; class controls how many R operations treat the object.

```{r coercion}
mixed_vector <- c(24, "unavailable")
print(mixed_vector)
stopifnot(is.character(mixed_vector))
```

**Expected:** `"24" "unavailable"`. Text coerces this whole atomic vector to character. Common atomic types are logical (`TRUE`), integer (`24L`), double (`24.5`), and character (`"S01"`). Conversion cannot recover an unknown measurement: `as.numeric("unavailable")` warns and produces `NA`. Keep explanations in a separate field.

## 2. Indexing, logic, and missingness

Positions start at one. Brackets select by position, name, or a logical vector. `&` and `|` compare elements; `!` negates a condition. Missing comparisons are unknown, so explicitly exclude missing values when selecting observed measurements.

```{r selections}
names(expression_values) <- sample_ids
print(expression_values[c(1, 4)])
print(expression_values["S05"])
measured <- !is.na(expression_values)
above_30 <- expression_values[measured & expression_values > 30]
print(above_30)
print(which(measured & expression_values > 30))
stopifnot(identical(unname(above_30), c(31, 37)))
stopifnot(length(expression_values[0]) == 0L, sum(!measured) == 1L)
```

**Expected:** positions 1 and 4 return `24, 28`; S05 returns `37`; selection above 30 returns `31, 37`, at positions `2, 5`. Index zero selects nothing. Use `is.na(x)`, never `x == NA`.

```{r missing-summary}
print(mean(expression_values))
print(mean(expression_values, na.rm = TRUE))
print(c(total = length(expression_values), measured = sum(measured), missing = sum(!measured)))
stopifnot(is.na(mean(expression_values)))
stopifnot(isTRUE(all.equal(mean(expression_values, na.rm = TRUE), 30)))
```

**Expected:** `NA`, then `30`; total 6, measured 5, missing 1. The available-case mean is 30 arbitrary units from five specimens. A missing specimen remains in the study denominator. Removing unavailable values from a summary does not explain why they were missing or eliminate selection bias.

## 3. Containers, factors, and observation units

A list holds heterogeneous components; `[ ]` retains a list and `[[ ]]` extracts one component. A matrix is rectangular with one shared type. A data frame can have differently typed columns of equal length.

```{r structures}
study_bundle <- list(gene = "Marker_A", expression = expression_values, observed = measured)
stopifnot(is.list(study_bundle["expression"]), is.numeric(study_bundle[["expression"]]))
expression_matrix <- rbind(Marker_A = c(24, 31, NA, 28, 37, 30),
                           Marker_B = c(8, 10, 9, 11, 7, 12))
colnames(expression_matrix) <- sample_ids
print(dim(expression_matrix))
expression_data <- data.frame(
  Sample_ID = sample_ids,
  Gene = rep("Marker_A", 6),
  Group = factor(rep(c("Reference", "Comparison"), 3),
                 levels = c("Reference", "Comparison")),
  Expression = unname(expression_values)
)
str(expression_data)
stopifnot(identical(dim(expression_matrix), c(2L, 6L)))
stopifnot(nrow(expression_data) == 6L, ncol(expression_data) == 4L)
stopifnot(!anyDuplicated(expression_data[c("Sample_ID", "Gene")]))
```

**Expected:** a 2-gene by 6-sample matrix and a six-row, four-column data frame. The matrix row is a gene and column is a specimen. In the long table, one row is one **gene–sample measurement**. A larger long table with two genes per specimen would have 12 rows but only six specimens; those rows are not 12 independent people. Sample identifiers need not identify people in every study.

Group is a fictional category assigned for practice, not a treatment assignment or diagnosis. Factor levels determine display order. `as.numeric(Group)` would give internal codes, not a biological measurement. Labels outside declared levels become missing; check them before conversion.

```{r rectangular-selection}
print(expression_data[1:2, c("Sample_ID", "Expression")])
one_column <- expression_data[, "Expression"]
one_column_table <- expression_data[, "Expression", drop = FALSE]
print(expression_matrix["Marker_B", c("S01", "S02")])
stopifnot(is.numeric(one_column), is.data.frame(one_column_table))
stopifnot(identical(dim(one_column_table), c(6L, 1L)))
```

**Expected:** S01 = 24, S02 = 31; Marker_B values for these specimens = 8 and 10. In `data[rows, columns]`, the comma separates axes. `drop = FALSE` preserves dimensions. Inspect `names()`, `rownames()`, `colnames()`, `str()`, and `dim()` before deciding which axis to summarize.

## 4. Decisions, loops, and checked functions

An `if` needs one nonmissing logical value. `&&` is a scalar short-circuit operator, useful for checking a scalar input before testing its value. A loop repeats an operation; `seq_along()` also behaves safely for an empty vector.

```{r decisions-and-loop}
if (sum(measured) >= 4L) {
  message("Enough observed values for this descriptive practice")
} else {
  message("Too few observed values for this practice rule")
}
centred_expression <- numeric(length(expression_values))
for (j in seq_along(expression_values)) {
  centred_expression[j] <- expression_values[j] - mean(expression_values, na.rm = TRUE)
}
print(centred_expression)
stopifnot(isTRUE(all.equal(unname(centred_expression), c(-6, 1, NA, -2, 7, 0))))
```

**Expected:** the first message and deviations `-6, 1, NA, -2, 7, 0` arbitrary units. A vectorized subtraction gives the same result. The threshold of four is a practice rule, not a clinically justified sample size.

```{r checked-function}
measured_mean <- function(values) {
  if (!is.numeric(values)) stop("values must be numeric")
  if (any(!is.finite(values[!is.na(values)]))) stop("observed values must be finite")
  if (all(is.na(values))) stop("at least one measurement is needed")
  mean(values, na.rm = TRUE)
}
print(measured_mean(expression_values))
rejected_text <- tryCatch(measured_mean("unknown"), error = function(e) conditionMessage(e))
rejected_empty <- tryCatch(measured_mean(c(NA_real_, NA_real_)), error = function(e) conditionMessage(e))
stopifnot(measured_mean(expression_values) == 30)
stopifnot(identical(rejected_text, "values must be numeric"))
stopifnot(identical(rejected_empty, "at least one measurement is needed"))
```

**Expected:** mean 30 and two caught validation errors. A function accepts inputs, makes a calculation, and returns a reusable value. `tryCatch()` here demonstrates errors deliberately; it does not turn invalid data into valid results. This function is numeric rather than expression-specific: a scientific workflow must separately validate the measurement definition and units.

## 5. Structured import and export

CSV uses comma-separated fields and may quote text; TSV uses tabs. The separator, header, missing-value convention, and row unit form an input contract. This temporary CSV is constructed entirely from the practice objects.

```{r table-round-trip}
expression_file <- tempfile(fileext = ".csv")
write.csv(expression_data, expression_file, row.names = FALSE, na = "NA")
reloaded <- read.csv(expression_file, na.strings = "NA", stringsAsFactors = FALSE)
unlink(expression_file)
stopifnot(identical(names(reloaded), names(expression_data)), nrow(reloaded) == 6L)
stopifnot(is.numeric(reloaded$Expression), sum(is.na(reloaded$Expression)) == 1L)
stopifnot(isTRUE(all.equal(reloaded$Expression, expression_data$Expression)))
print(sapply(reloaded, class))
```

**Expected:** Sample_ID, Gene, and Group import as character; Expression is numeric. CSV does not preserve factor levels. `row.names = FALSE` prevents an extra index column. In real work, retain raw files and their dictionary, validate unique gene–sample pairs, and record units and software versions before analysis.

## 6. A labeled distribution

A histogram counts measurements in intervals. Its horizontal units and vertical denominator must be explicit. It describes this invented set, not an expression assay's biological distribution.

```{r expression-histogram}
expression_hist <- hist(expression_data$Expression, breaks = c(20, 25, 30, 35, 40),
                        main = "Marker_A: five synthetic observed specimens",
                        xlab = "Expression (arbitrary simulated units)", ylab = "Observed specimens",
                        col = "lightblue", right = FALSE)
print(expression_hist$counts)
stopifnot(identical(expression_hist$counts, c(1L, 1L, 2L, 1L)))
```

**Expected:** counts `1, 1, 2, 1`, totaling five. Left-closed bins place 30 in the bin starting at 30. Different breaks change appearance, not the measurements. Five observations cannot establish a population distribution, a treatment effect, or differential expression.

## Common mistakes and debugging

| Symptom | Cause and correction |
| --- | --- |
| Object not found | Run its defining block and check case-sensitive spelling |
| Arithmetic on text | Inspect types; separate invalid strings from missing numeric values |
| Mean is NA | Count missingness and choose a documented available-case rule |
| Wrong observations selected | Confirm gene/sample axes and the row–column comma |
| Table becomes a vector | Preserve dimensions with `drop = FALSE` |
| Invalid conditional | Reduce a vector to one decision, or use elementwise operations |
| Import has an extra column | Check header/separator and whether row names were exported |

## Independent practice

1. Construct an eight-specimen table for an invented gene with two unavailable values. Report total and measured counts and select above a stated arbitrary-unit threshold.
2. Create a two-gene matrix for those specimens. Explain why it contains 16 gene–sample cells rather than 16 people.
3. Test `measured_mean()` with a finite vector, text, all missing values, and infinity. Predict each outcome first.
4. Export/import your table and recreate Group's levels explicitly. Validate identifiers and values.
5. Draw two histogram break sets; explain what changes and what remains fixed.

## Ready to move on

Keep a runnable script and a short data dictionary. Restart R and rerun everything before interpreting the outputs.

| Evidence | Mastery check |
| --- | --- |
| Objects | Inspect and explain all five containers and coercion |
| Selection | Select gene/sample fields and observed values using logic |
| Control flow | Trace one conditional, loop, and checked function |
| Import/export | Recover six rows and numeric values; explain lost factor metadata |
| Scientific interpretation | Mean 30 arbitrary units from 5 of 6 specimens; distinguish gene, sample, and observation unit |

## Further learning

[Next: Week 2](Week_2_Group_Work_Narrative_Walkthrough.md) · [Weekly index](README.md) · [R function sheet](../06_RESOURCES/R/EASY_FUNCTION_SHEET.md) · [Paul's Notes](../README.md)

References: [R extraction](https://stat.ethz.ch/R-manual/R-devel/library/base/html/Extract.html), [data import/export manual](https://cran.r-project.org/doc/manuals/r-release/R-data.html), and installed help `?factor`, `?hist`, `?getwd`.

## Weekly Learning Objectives & Mastery Assessment

Assess your own new synthetic specimen example after completing the lesson. These criteria describe this independent study guide; they are not institutional marking criteria.

| Measurable objective | Evidence of competency | Independent mastery criterion |
| --- | --- | --- |
| Choose and inspect R containers | A script using a vector, list, matrix, factor, and data frame; recorded type, class, and dimensions | Explain each choice and predict the result of one coercion before running it |
| Select observations without losing structure or missingness | Positional, named, and logical selections with printed identifiers and measured/missing counts | Reconcile selected IDs against the input and preserve table dimensions when selecting one column |
| Implement checked computation | One function plus a conditional and loop; valid and invalid input trials | Predict each return or error, reject invalid values, and handle an all-missing input deliberately |
| Recover a tabular export | CSV round trip, data dictionary, and explicit reconstruction of category levels | Verify identifiers and numeric values; explain which factor metadata the CSV does not retain |
| Interpret a distribution | Two labeled histograms and a measured-count summary with units | Explain how breaks affect the display, distinguish specimens from gene–sample cells, and make no population or clinical claim from the synthetic values |

**Mastery decision:** meet every row using fresh invented inputs and a clean-session rerun, with your own script and explanation rather than copied outputs. Keep predictions, observed checks, and any corrections. If a criterion fails, revisit that topic and demonstrate it again before progressing.
