---
title: "HSE 711 — Week 1: Introduction to R Lecture Notes"
author: "Independent course study companion"
output:
  html_document:
    toc: true
    toc_depth: 2
    toc_float:
      collapsed: false
    code_folding: show
    css: ../06_RESOURCES/Presentation/reading.css
---

## Purpose and learning objectives

These notes explain **how to approach** the first Introduction to R assignment using the methods introduced in the Week 1 lecture: object classes, vectors and lists, indexing, logical comparisons, loops, data frames, file input, graphics, and functions. They provide **parallel practice examples**, not the assignment answers, submission, or supplied dataset. The original lecture source is referenced by the existing Week 1 group-work companion but is not reproduced here.

By the end, you should be able to explain the reasoning for every line of your own code, inspect results, and independently adapt the techniques to the assigned inputs.

| Assignment task | Lecture technique to revisit | Check your understanding |
| --- | --- | --- |
| 1a–b: identify and count object types | `list()`, `class()`, `for`, `table()` | Why inspect each list component separately? |
| 1c–e: comparisons, factors, positions | `is.numeric()`, `if`, `factor()`, `which()` | How do you avoid comparing letters with numbers? |
| 2a–c: import and dimensions | `read.delim()`, `row.names`, `nrow()`, `ncol()` | Which axis represents genes versus samples? |
| 2d–e: graph and export | `hist()`, `[rows, columns]`, `write.csv()` | What happens when a matrix or table is subset incorrectly? |
| 3: categorize values with a function | `function()`, `for`, `if/else`, `return()` | What does the function accept and produce? |

## 1. Understanding R objects before classifying them

**Lecture idea:** R stores values inside named objects. `<-` assigns a value; `class()` describes an object's class. A list can contain different classes in separate elements, unlike an atomic vector.

```{r week1-types}
practice_list <- list(12, "alpha", TRUE, 2.5)

# Inspect one component; double brackets extract its contents.
class(practice_list[[1]])
class(practice_list[[2]])

# Revisit the loop syntax from the lecture.
practice_types <- character(length(practice_list))
for (i in seq_along(practice_list)) {
  practice_types[i] <- class(practice_list[[i]])
}
practice_types
table(practice_types)
```

**Why it works:** `seq_along()` supplies the positions of the list; `[[i]]` retrieves each underlying value; `class()` converts the observed class into a label. `table()` counts repeated labels. An alternative is `vapply(practice_list, class, character(1))` when each element has one class label.

**Checkpoint:** Explain why `class(practice_list)` alone is not an answer to the question about *each component*.

## 2. Comparing numbers safely, making factors, and locating positions

A comparison like `value > threshold` only makes sense if the value has a compatible type. An `if` statement executes only when the condition is a single TRUE or FALSE. `&&` is appropriate for a scalar short-circuit condition inside this loop.

```{r week1-conditions}
mixed_list <- list("north", 3, 11, "south", 7)
n_large <- 0L

for (value in mixed_list) {
  if (is.numeric(value) && length(value) == 1L &&
      !is.na(value) && value > 6) {
    n_large <- n_large + 1L
  }
}
n_large

# Categorical classes can be represented with a factor.
practice_factor <- factor(vapply(mixed_list, class, character(1)))
levels(practice_factor)

# which() gives one-based positions, not the matching values.
positions <- which(vapply(
  mixed_list,
  function(value) is.numeric(value) && length(value) == 1L &&
    !is.na(value) && value == 7,
  logical(1)
))
positions
```

**Why it works:** The type check protects the numeric comparison. `factor()` constructs category levels; `which()` converts TRUE entries of a logical vector to indices. R indices start at **1**, not 0.

**Checkpoint:** Change the search target and predict whether the result will contain zero, one, or multiple indices.

## 3. Importing a gene-by-sample expression table

**Lecture idea:** Inspect a data frame's structure *before* processing it. `read.delim()` reads tab-separated files. `header = TRUE` uses the top row for column labels; `row.names = 1` uses the first column for row identifiers. These are different options.

The illustrative code below is intentionally **not run**: the exercise's original clinical trial file is an independently supplied classroom input, not a public repository asset.

```{r week1-import, eval=FALSE}
input_path <- "data/your_local_expression_table.txt"
stopifnot(file.exists(input_path))

expression_df <- read.delim(
  file = input_path,
  header = TRUE,
  row.names = 1,
  check.names = FALSE
)

dim(expression_df)             # rows, columns together
nrow(expression_df)            # number of gene rows
ncol(expression_df)            # number of sample columns
head(rownames(expression_df))  # gene identifiers
head(colnames(expression_df))  # sample identifiers
str(expression_df)             # storage types
```

**Why it works:** Each gene is an observation along the row axis and each sample occupies a column in this file layout. `check.names = FALSE` avoids automatic renaming of sample identifiers. `head()` and `str()` help confirm the file was interpreted correctly.

**Checkpoint:** A table is sometimes oriented with samples in rows. How would you establish the actual orientation instead of assuming it?

## 4. Choosing one gene and reading its histogram

A histogram describes the **distribution of numeric values** for one selected gene across the samples. It displays how many values fall inside intervals; it does not identify patient-level clinical outcomes or establish treatment effects.

```{r week1-histogram}
# Small demonstration table, unrelated to the assigned trial.
toy_expression <- data.frame(
  sample_a = c(3, 12, 6),
  sample_b = c(7, 14, 9),
  sample_c = c(5, 11, 8),
  sample_d = c(6, 13, 7),
  row.names = c("gene_a", "gene_b", "gene_c")
)

gene_id <- "gene_b"
gene_values <- as.numeric(toy_expression[gene_id, ])

hist(
  gene_values,
  breaks = 3,
  main = paste("Distribution for", gene_id),
  xlab = "Expression value (illustrative units)",
  ylab = "Number of samples"
)
```

**Read the plot:** The horizontal axis is the measured expression value; the vertical axis counts samples in each interval. The number of bins affects appearance. A long tail suggests a few comparatively high observations, but a histogram alone does not explain their cause.

**Checkpoint:** Why must the selected row become a numeric vector before calling `hist()`?

## 5. Selecting rows and columns, then exporting carefully

R's two-dimensional syntax is `data[rows, columns]`. Always inspect the expected resulting dimensions before writing a file. `drop = FALSE` retains a table-like object even when one row or one column is selected.

```{r week1-subset}
toy_sub <- toy_expression[1:2, 1:2, drop = FALSE]
dim(toy_sub)
rownames(toy_sub)
colnames(toy_sub)

# To export an independently created practice table:
# write.csv(toy_sub, file = "toy_sub.csv", row.names = TRUE)
```

**Checkpoint:** Why is `[1:2, 1:2]` not equivalent to `[1:2]`? How do row identifiers appear in the exported CSV?

## 6. Writing a reusable categorization function

**Lecture idea:** A function accepts an argument, performs the same logic consistently, and returns a result. The `if/else` branches belong **inside** the function, and the function is called only after it has been defined.

```{r week1-function}
group_measurements <- function(values, cutoff = 10) {
  lower <- numeric(0)
  higher <- numeric(0)

  for (value in values) {
    if (is.na(value)) {
      next
    } else if (value <= cutoff) {
      lower <- c(lower, value)
    } else {
      higher <- c(higher, value)
    }
  }

  list(
    at_or_below = sort(lower),
    above = sort(higher)
  )
}

practice_values <- c(14, 3, 8, 12, 5)
practice_result <- group_measurements(practice_values)
str(practice_result)
```

**Why it works:** The function parameter `values` is local to the function. `cutoff` has a default, so the same method can be tried with a different threshold. The returned list preserves both categories for later use. `next` explicitly skips missing values; for real analyses, the decision to omit missing values must be justified.

**Checkpoint:** Predict the output before changing the cutoff. What needs to change if the input contains strings instead of numeric values?

## 7. Explaining your method, not just showing output

For each part of the assignment, write one or two sentences answering: **What object did I start with? What did this function, loop, or index operation do? How did I check the result?** Distinguish a descriptive graph from an inferential result.

A useful repeatable workflow is: inspect the data, choose the operation, run a small test, verify dimensions or positions, and only then apply the method to the full input. Keep your own solutions and any classroom-restricted data in the appropriate private or submission location.

## 8. Final self-check

- Can I explain `class()`, `table()`, `factor()`, `which()`, and why the list requires component-wise inspection?
- Can I describe why `row.names = 1` is distinct from `header = TRUE`?
- Can I identify the gene and sample axes and confirm an export's dimensions?
- Can I explain each argument to `hist()` without claiming an unsupported biological finding?
- Can I trace a function call into its loop, branches, and return value?

**Related:** [Week 1 group-work walkthrough](../03_Group_Work/Week_1_Group_Work_Narrative_Walkthrough.md) · [R function sheet](../06_RESOURCES/R/EASY_FUNCTION_SHEET.md) · [Lecture index](README.md)
